Review of the Literature on Exact Formulas for Mediation Effects

### 2024-05-16 notes/Literature Review/Literature Review.tex
Expand lit review. Add Imai group and assorted miscellany

@@

@@

\subsection{What to do Next}

@@

\section{The \citeauthor{Sam23} group}
\label{sec:Sam}

@@

@@

Defines mediation effects as partial derivatives. Does UQ by bootstrap or by a very rough method called MCCI. This MCCI simulates parameter values from a normal distribution parameterized by the estimates, then evaluates the mediation effects on these simulated parameters. This gives a distribution, but it seems to me that they would be better off just using the $\delta$-method.

@@

There does seem to be some work on applying mixed-effects methodology to SEMs. I'm not sure yet if this has made its way to the SEM-based mediation analysis literature. I'll look into this next week.

@@

@@

@@

@@

>>>>

\usepackage{hyperref}   % For the \url{} command

@@

\subsection{May 16}

This week, I added work by Imai and colleagues, as well as a few general references. See Section \ref{sec:Imai} for the Imai group. There is also a method proposed by \citet{Lan12} based on inverse probability weighting to estimate unobserved counterfactuals. Both of these methods are included in the extensive Monte Carlo study of \citet{Sam23} and found to be comparable to inference for mediation effects based on the $\delta$-method. However, the method of \citet{Ima10I} is computationally intensive, and that of \citet{Lan12} is quite different from our approach of combining estimates from two regression models.

I also found an \texttt{R} package, \texttt{merDeriv} \citep{Wan18} which computes the full asymptotic covariance matrix for a GLMM fit using \texttt{lme4}. This will save me a lot of time in implementing the $\delta$-method uncertainty quantification for mixed-effects models.

\subsubsection{What to do next}

I'm still not totally confident about the SEM literature, but otherwise I don't think anyone has done exact asymptotic theory for multilevel mediation analysis. I feel ready to start working on the mixed-effects version of our analysis. I will continue reading, but I think it makes sense to start spending some time writing code.

There are some extensions to the methodology outlined by B \& B that I think we should consider. First, we will want to include direct and indirect effects in addition to our total effects. We can follow \citet{Sam23} for this, although we will also need to give some though to how we incorporate random effects into the definitions. It would also be nice to explicitly estimate the ``mediation proportion''. For continuous response, this is the indirect effect divided by the total effect. It's less clear to me what definition to use for multiplicative effects (e.g. effects on odds-ratio scale).

Second, it is common to include an interaction effect between the exposure and mediator in our model for the outcome\footnote{There is no interaction effect in our trust study because a preliminary analysis found that it didn't have a significant effect on the outcome.}. This isn't conceptually any harder, it just adds another component to our formulas.

I think that we could address just the above two points and, with some nice Monte Carlo studies, call it a paper. This would resemble the work by \citeauthor{Sam23} (e.g. \citealp{Sam23}). After that, there are some more dramatic extensions we can consider.

One extension that we really should look at is predicted cluster-specific effects. The uncertainty quantification here is subtle; see, e.g., \citet{Skr09,Boo98}. The Imai group does give cluster-specific mediation effects, but only under their restricted framework. Otherwise, this seems to still be an open problem

Finally (for now), I think we should eventually look at sensitivity analysis. There are a bunch of assumptions required for us to give a causal interpretation to our results. My understanding is that sensitivity analysis lets us put a bound on the extent to which these assumptions can be violated without invalidating our conclusions. I'm still looking into how this works.

\subsection{May 2}

@@

\subsubsection{What to do Next}

@@

\section{The Samoilenko and Lefebvre group}
\label{sec:Sam}

@@

Compares their approach to those of the Imai group (see Section \ref{sec:Imai}), as well as \citet{Lan12}. Found that the $\delta$-method based approach proposed by the \citeauthor{Sam23} group gives comparable performance with its computationally intensive alternatives.

@@

Defines mediation effects as partial derivatives. Does UQ by bootstrap or by a very rough method called MCCI. This MCCI simulates parameter values from a normal distribution parameterized by the estimates, then evaluates the mediation effects on these simulated parameters. This is the same UQ method used by the Imai group (see Section \ref{sec:Imai}).

@@

There does seem to be some work on applying mixed-effects methodology to SEMs. See, e.g., \citet{Rab04} for early work in this direction.

I'm not sure yet if this has made its way to the SEM-based mediation analysis literature. I'll look into this next week.

@@

\section{The Imai Group}
\label{sec:Imai}

This group focuses mainly on the assumptions required to link regression-based mediation analysis with causal inference. They give definitions of mediation effects which closely resemble the ``mediation formula'' \citep{Pea12}, but only for differences of conditional expectation (i.e. the ``risk-difference scale''; see also \citealp{Sam23}). Their approach to UQ is very computational. They recommend either the bootstrap or a simulation-based approach outlined in \citet{Kin00}. Briefly, simulate parameters from their fitted sampling distributions, then evaluate the quantity of interest on each sample and observe the resulting distribution. I think this is connected to empirical Bayes methodology, but I haven't looked into it extensively.

Their methods are implemented in the \texttt{R} package \texttt{mediation} \citep{Tin14}. They can handle mixed-effects models, including prediction of group-level effects.

\subsection{\citet{Ima10I}}

This paper presents an UQ algorithm for estimating mediation effects defined as differences of expected values. The algorithm closely resembles the bootstrap, but is slightly different. Specifically, parameters are repeatedly sampled from the fitted sampling distributions of their estimators, then mediation effects are estimated using the sampled parameter estimates \citep[see, e.g.,][]{Kin00}. They also give an alternative method which just uses the bootstrap.

Refers to the causal inference theory from \citet{Ima10II} to justify identification of counterfactuals with functions of conditional expectations.

\subsection{\citet{Ima10II}}

Focuses on estimating the indirect effect for a continuous response. Defines this effect as a difference of counterfactuals and identifies this difference with a conditional expectation in the same manner as the mediation formula given by \citet{Pea12}.

Also gives a ``non-parametric estimator'' for categorical mediator based on a contingency table based approach. The idea is to partition the data based on the mediator, estimate the mediation effect within each group with a difference of averages, then re-combine by averaging.

UQ is given either by the $\delta$-method or by the bootstrap. Some discussion of sensitivity analysis.

\subsection{\citet{Ima11}}

Very similar to \citet{Ima10I} and \citet{Ima10II}. Focuses more on causal inference interpretation and applications to political science. Doesn't appear to advance the methodology.

\subsection{\citet{Tin14}}

This paper overviews the \texttt{mediation} package, which implements the methods developed by this group. Can handle GLMMs, but only operates on differences in the expected value of the outcome (i.e. risk-difference scale for binary outcome). Does UQ for parameters \textbf{and predicted random effects}, but doesn't go into detail on how. I'm guessing that it just follows the Monte Carlo method given by \citet{Ima10I}.

Package last updated 2019-10-08

\section{The \texttt{SPSS} Macro \texttt{MLmed}}
\label{sec:MLmed}

This macro appears to be the standard way to do multilevel mediation analysis in \texttt{SPSS}. As such, it is likely to be used extensively by applied researchers, and we should understand it. The macro can only handle continuous outcomes and mediators.

\subsection{\citet{Roc19}}

User manual for the \texttt{MLmed} macro. Can only handle continuous outcomes and mediators. UQ is either by the bootstrap or the ``Monte Carlo method''. See \citet{Pre12} or \citet{Wil08}.

\subsection{\citet{Hay20}}

A recommended reference for the methodology used in the macro.

\subsection{\citet{Roc17}}

A conference poster presenting the macro. Popular reference for people looking to cite the macro.

\section{Applications}

It doesn't hurt for us to point to what people are doing in the applied literature. Here is some of what shows up when I search ``multilevel mediation analysis'' on Google Scholar.

\subsection{\citet{Vea19}}

Does multilevel mediation analysis with continuous outcome and mediator. UQ is done with ``Monte Carlo confidence intervals''. Uses an \texttt{SPSS} macro called \texttt{MLmed}. See Section \ref{sec:MLmed}.

\subsection{\citet{Ara21}}

It looks like their variables are all ordinal (1-6 or 1-8 Likert scale). I'm not sure how they fit this into the mediation analysis framework. Moving on, they use \texttt{Mplus} to do their analysis via ``maximum likelihood''. Not much detail given.

\subsection{\citet{Hwa12}}

Follows an earlier edition of \citet{Mac17}, as well as the very early work of \citet{Bar86}. As in \citet{Ara21}, variables are ordinal/Likert. Analysis is done in \texttt{SPSS}, although few details are given. Note that this work was published before the \texttt{SPSS} macro of \citet{Roc19}.

@@

\subsection{\citet{Kru01}}

Throughout some parts of the literature on multilevel mediation analysis, you will see people talk about, e.g., the 2-1-1 model. This means that the exposure is measured at ``level 2'' (group-level), while the mediator and outcome are measured at ``level 1'' (individual-level). That is, in this model we would include a random effect for the mediator, but not the exposure. Other examples of such models include 2-2-1 and 1-1-1. As far as I can tell, this notation was proposed by \citet{Kru01}.

This paper also does $\delta$-method uncertainty quantification for the very simple mediation effects that people were using at that time (i.e. a product of coefficients). In fact, this approach for testing the presence/absence of mediation effects dates back to \citet{Sob82}.

@@

\subsection{\citet{Lan12}}

Proposes estimating mediation effects using inverse probability weighting. Duplicates the observed data and uses weighting to estimate unobserved counterfactual outcomes. Referred to as the ``weighting method''. Does give analytical (asymptotic) standard errors based on the sandwich variance estimator that arises from the theory of generalized estimating equations. See their ``Web Appendix 3'' for more details. They also suggest using the bootstrap.

Implemented in the \texttt{R} package \texttt{medflex} \citep{Ste17}.

@@

\subsection{\citet{Wan18}}

Standard reference for the \texttt{merDeriv} package in \texttt{R}. This is a small package that computes the Hessian of the observed data likelihood with respect to all parameters (including RE covariance parameters) for LMMs and GLMMs fit using \texttt{lme4}. In particular, use the \texttt{vcov} function with \texttt{full=TRUE} to get the asymptotic covariance matrix of the parameters (i.e. the inverse of the negative Hessian of the log-likelihood). You can also choose your parameterization for the RE parameters from variances, SDs or terms in the Cholesky decomposition. The last one is referred to as \texttt{theta} in \texttt{lme4} and is used extensively therein.
